On the presentation of orthogonal groups over polynomial rings

### 2019-01-30 dl-paper.tex
Reorganizing introduction

\subsection{Unstable Quillen K-groups}
\begin{comment}
In order to better understand $C(\Phi, R, I)$  we would like to continue the sequence~\eqref{eq:relative-St} to the left. 
Thus we need to work with unstable $\K_3$-functors a la Quillen (cf.~\cite{Pa89}).
\end{comment}
Throughout this section $G, P, S$ denote three group-valued functors on the category of commutative rings satisfying the following assumptions:
\begin{enumerate}
 \item\label{req:left-exact} $G$ is left exact;
 \item $P$ is a subfunctor of $G$ such that for all $R$ the group $P(R)$ is a perfect normal subgroup of $G(R)$;
 \item There is a natural transformation $\varphi \colon S \to P$ such that $ \varphi \colon S(R) \to P(R)$ is a universal central extension for all $R$;
 \item \label{req:coeq} $S$ preserves coequalizers of kernel pairs.
\end{enumerate}

Thanks to the functoriality of Quillen's plus-construction (see e.\,g.~\cite[Proposition~5.2.4]{Ro95}) one can define a functor 
$X_{G, P}(R) = BG(R)^+_{P(R)} \colon \catname{CRings} \to \catname{Top}$ and then define for $i \geq 1$ the corresponding K-group functors $\K_{i}(G, P, R)$ as the homotopy groups $\pi_i(X_{G, P}(R))$.

Let us also briefly recall Loday's procedure for relativization of $S$, that is,
 extending $S$ from the category of rings to the category of pairs.
Denote by $p_0, p_1, \Delta$ the two obvious projections and the diagonal map $\xymatrix{D_{R,I} \ar@<1pt>[r] \ar@<-2.5pt>[r] & \ar@<-4pt>[l] R}$
  and set $d_i = Sp_i$, $s = S\Delta$.
 Obviously, $d_0s = d_1 s = id_{S(R)}$. Set $G_i = \Ker(d_i)$, and define $S(R, I)$ as $ G_0 / [G_0, G_1]$, denote by $\mu$ the natural map $S(R, I) \to S(R)$ induced by $d_1$.

\begin{prop}\label{characterization}
\begin{proplist}
 \item There are natural isomorphisms
  \[ \K_2(G, P, R) \cong \Ker(S(R) \to G(R)), \ \K_3(G, P, R) \cong \HH_3(S(R), \ZZ). \]
 \item The map $\mu \colon S(R, I) \to S(R)$ is a universal relative central extension of 
    $S(R) \to S(R/I)$, whose kernel is naturally isomorphic to the third relative homology group $\HH_3(S(R), S(R/I), \ZZ)$.
 \item There is an exact sequence natural in $(R, I)$:
 \[ \xymatrix{\K_3(G, P, R) \ar[r] & \K_3(G, P, R/I) \ar@{->>}[r] & \Ker(\mu).} \]
 \end{proplist}   
\end{prop}
\begin{proof}
 The first claim follows from the solution of Exercises~1.8--1.9 of \cite[Chapter~IV]{Kbook}. %provided one verifies that all the intermediate isomorphisms are natural in $R$.

 Let us verify the second claim. From~\eqref{req:left-exact} and $\varphi(G_i) \subseteq \Ker(G(p_i))$ we conclude that $G_0 \cap G_1 \subseteq \Ker(\varphi \colon S(D_{R, I}) \to G(D_{R, I}))$
  and hence $G_0 \cap G_1$ is a central subgroup of $S(D_{R, I})$.
 Now invoking~\cite[Proposition~6]{Lo78} we conclude that the extension $S(R, I) \to S(R)$ is a universal relative central extension of the coequalizer $\mathrm{coeq}(d_0, d_1)$, 
   which coincides with $S(R) \to S(R/I)$ by~\eqref{req:coeq}. The kernel of $\mu$ is isomorphic to $\HH_3(S(R), S(R/I), \ZZ)$ by~\cite[Th{\'e}or{\`e}me~2]{Lo78}.

 The third claim now follows from the starting portion of the homology long exact sequence of the pair $(S(R/I), S(R))$ and the first two assertions.
\end{proof}

Stability for orthogonal and symplectic K-theories is given by the following theorem of Panin
\begin{theorem}[Panin] \label{lem:Panin-stability}
 Let $R$ be either a field, principal ideal domain or a Dedekind domain and set $a = 1,2,3$ in each of respective cases.
 Then the canonical map $\K_k(\rD_\ell, R) \to \K_k(\rD_\infty, R)$ is an epimorphism for $\ell \geq \mathrm{max}(2k, a+k-1)$ 
 and an isomorphism for $\ell \geq \mathrm{max}(2k+1, a+k)$, where $a=1,2,3$ in the case when $R$ is a field, p.i.d. or a Dedekind domain, respectively \end{theorem}
\begin{proof} See~\cite[Theorem~9.4]{Pa89}. \end{proof}

\begin{comment}
 Stabilization for orthogonal and symplectic K-groups a la Quillen is given by Panin's theorem,  More specifically the bounds for surjective (resp. injective) stability is 
 $\ell \geq \max(6, a+2)$ (resp. $\ell \geq \max(7,a+3)$) for $a=1,2,3$ in the case when $R$ is a field, p.i.d. or a Dedekind domain, respectively.
\end{comment}

\begin{comment}
Suppose for a moment that $\langle \alpha, \beta \rangle = -1$ and  $\langle \beta, \alpha \rangle = -1$ then
\[ \{s, t^{-1} \} = \{s,  t^{-1}\}_\alpha = \{t, s^{-1} \}_\beta^{-1} = \{s^{-1}, t\} \]
In particular, $\{s, s^{-1}\} = \{s, s^{-1}\}^{-1}$ 
\end{comment}

@@

%\subsection{Isogenous forms of the orthogonal group}
\subsection{Hermitian K-theory and higher Grothendieck--Witt groups}
In this section we collect some facts pertaining to Hermitian K-theory that will be of relevance to us. We refer the reader to~\cite[\S~2]{AF17} or~\cite[\S~2]{FRS12} for a more detailed introduction.

First of all, recall that in~\cite{Ka80} M.~Karoubi defines higher Grothendieck--Witt groups via Quillen's plus-construction:

Schlichting gave a more general definition which still agrees with the classical one in the affine case.
In particular one has:
\begin{equation}
 \GW_i^{[k]}(R) = \left\{\begin{array}{ll} \mathrm{KO}_i(R), & k = 0, \\ U_i(R), & k = 1, \\ \mathrm{KSp}_i(R), & k = 2, \\ {}_{-1}\!U_i(R), & k = 3. \end{array}\right.
\end{equation}

\begin{theorem}[Bass Fundamental Theorem]\label{bass-ft} Suppose that $2 \in R^*$, then for any $i\geq 1$, $k\in \ZZ/4\ZZ$ there is a split exact sequence of abelian groups
 \[ \xymatrix{ 0 \ar[r] & \GW_i^{[k]}(R) \ar[r] &  \GW_i^{[k]}(R[X, X^{-1}]) \ar[r]  & \GW_{i-1}^{[k-1]}(R) \ar[r] & 0.} \] \end{theorem}
\begin{proof} This is a special case of~\cite[Theorem~9.13]{Sch16} and the remark that follows it. \end{proof}

>>>>

\begin{comment}
Suppose for a moment that $\langle \alpha, \beta \rangle = -1$ and  $\langle \beta, \alpha \rangle = -1$ then
\[ \{s, t^{-1} \} = \{s,  t^{-1}\}_\alpha = \{t, s^{-1} \}_\beta^{-1} = \{s^{-1}, t\} \]
In particular, $\{s, s^{-1}\} = \{s, s^{-1}\}^{-1}$ 
\end{comment}
\subsection{Unstable Quillen K-groups}
\begin{comment}
In order to better understand $C(\Phi, R, I)$  we would like to continue the sequence~\eqref{eq:relative-St} to the left. 
Thus we need to work with unstable $\K_3$-functors a la Quillen (cf.~\cite{Pa89}).
\end{comment}
Throughout this section $G, P, S$ denote three group-valued functors on the category of commutative rings satisfying the following reqirements:
\begin{enumerate}
 \item \label{req:left-exact} $G$ is left exact;
 \item \label{req:subfunc} $P$ is a subfunctor of $G$ such that for all $R$ the group $P(R)$ is a perfect normal subgroup of $G(R)$;
 \item \label{req:uce} There is a natural transformation $\varphi \colon S \to P$ such that $ \varphi \colon S(R) \to P(R)$ is a universal central extension for all $R$;
 \item \label{req:coeq} $S$ preserves coequalizers of kernel pairs.
\end{enumerate}

Thanks to the functoriality of Quillen's plus-construction (see e.\,g.~\cite[Proposition~5.2.4]{Ro95}) one can define a functor 
%TODO: can also refer to Weibel's IV.1.5 provided Top is replaced with hTop which apparently does not change anything
$X_{G, P}(R) = BG(R)^+_{P(R)} \colon \catname{CRings} \to \catname{Top}$ and then define for $i \geq 1$ the corresponding K-group functors $\K_{i}(G, P, R)$ as the homotopy groups $\pi_i(X_{G, P}(R))$.

Let us also briefly recall Loday's procedure for relativization of $S$, that is,
 a way of extending $S$ from the category of rings to the category of pairs.
Denote by $p_0, p_1, \Delta$ the two obvious projections and the diagonal map $\xymatrix{D_{R,I} \ar@<1pt>[r] \ar@<-2.5pt>[r] & \ar@<-4pt>[l] R}$
  and set $d_i = Sp_i$, $s = S\Delta$.
 Obviously, $d_0s = d_1 s = id_{S(R)}$. Set $G_i = \Ker(d_i)$, and define $S(R, I)$ as $ G_0 / [G_0, G_1]$, denote by $\mu$ the natural map $S(R, I) \to S(R)$ induced by $d_1$.

%TODO: Give the definition of $D_{R, I}$ and of 
\begin{prop}\label{characterization}
\begin{enumerate}
 \item There are natural isomorphisms
  \[ \K_2(G, P, R) \cong \Ker(S(R) \to G(R)), \ \K_3(G, P, R) \cong \HH_3(S(R), \ZZ). \]
 \item The map $\mu \colon S(R, I) \to S(R)$ is a universal relative central extension of 
    $S(R) \to S(R/I)$, whose kernel is naturally isomorphic to the third relative homology group $\HH_3(S(R), S(R/I), \ZZ)$.
 \item There is an exact sequence which is natural in $(R, I)$:
 \[ \xymatrix{\K_3(G, P, R) \ar[r] & \K_3(G, P, R/I) \ar@{->>}[r] & \Ker(\mu).} \]
 \end{enumerate}   
\end{prop}
\begin{proof}
 The first claim follows from Exercises~1.8--1.9 of \cite[Chapter~IV]{Kbook} (here we use~\eqref{req:subfunc}--\eqref{req:uce}). %provided one verifies that all the intermediate isomorphisms are natural in $R$.

 Let us verify the second claim. From~\eqref{req:left-exact} and $\varphi(G_i) \subseteq \Ker(G(p_i))$ we conclude that $G_0 \cap G_1 \subseteq \Ker(\varphi \colon S(D_{R, I}) \to G(D_{R, I}))$
  therefore $G_0 \cap G_1$ is a central subgroup of $S(D_{R, I})$.
 From~\eqref{req:uce} we also obtain that $H_1(S(D_{R,I})) = H_2(S(D_{R,I})) = 0$ .
 The last two assertions allow us to invoke~\cite[Proposition~6]{Lo78} and conclude that the extension $S(R, I) \to S(R)$ is a universal relative central extension of the coequalizer $\mathrm{coeq}(d_0, d_1)$
  in the sense of~\cite[\S~3]{Lo78}. 
 Notice that this coequalizer coincides with $S(R) \to S(R/I)$ by~\eqref{req:coeq}. 
 Finally, the kernel of $\mu$ is isomorphic to $\HH_3(S(R), S(R/I), \ZZ)$ by~\cite[Th{\'e}or{\`e}me~2]{Lo78}.

 The third claim now follows from the starting portion of the homology long exact sequence of the map $S(R) \to S(R/I)$ and the first two assertions.
\end{proof}

Let us mention the example of a triple $(G,P,S)$ satisfying the above conditions.
\begin{example} \label{ex:orth}
 For $n \geq 5$ take the orthogonal group functor $\mathrm{O}_{2n}(-)$ and its elementary subfunctor $\mathrm{EO}_{2n}(-)$ as functors $G_n$ and $P_n$, respectively.
 Now set $S_n = \St(\rD_n, -)$.
 From [Stein's paper] and [Lemma on centrality] it follows that $S_n$ and $P_n$ satisfy~\eqref{req:uce}.
 The other requirements can be verified in a straightforward manner.
\end{example}

In the above definition we could have replaced the orthogonal group with a different isogeneous form of the orthogonal group, e.\,g. 
 we could have chosen $G_n = \mathrm{SO}_{2n}$ or $G_n = \mathrm{Spin}_{2n}$ and then defined $P_n$ as its elementary subfunctor.
Of course, this would change the functors $\K_i(G_n, P_n, -)$ for $i=1,2$ however, as the proof of~\cref{characterization} suggests
 it would not change the functors $\K_i(G_n, P_n, -)$ for $i\geq 3$.

Now let $G_n, P_n$ be again as in~\cref{ex:orth}. 
Let us formulate a stability result for orthogonal Quillen K-groups (see~\cite[Theorem~9.4]{Pa89}).
\begin{theorem}[Panin] \label{lem:Panin-stability}
 Let $R$ be either a field, principal ideal domain or a Dedekind domain. Set $a = 1,2$ or $3$ in each of these respective cases.
 Then the canonical map $\K_i(G_n, P_n, R) \to \K_i(G_{n+1}, P_{n+1}, R)$ is an epimorphism for $n \geq b$ 
 and an isomorphism for $\ell \geq b + 1$, where $b = \mathrm{max}(2i, a+i-1)$. \end{theorem}

Of course, in the limit, the groups $\K_i(\mathrm{O}_\infty, \mathrm{EO}_\infty, R) = KO_i(R)$ coincide with Karoubi's hermitian groups ${}_{-1}\!L_i(R)$ defined in~\cite{Ka80}.
The modern version of hermitian K-theory is the theory of higher Grothendieck--Witt groups of schemes.
We refer the reader to~\cite[\S~2]{AF17} or~\cite[\S~2]{FRS12} for a more detailed introduction to it.
For our purposes it would be sufficient to think of $\GW_i^{[k]}(R)$ as a shorthand for the following:
\begin{equation}
 \GW_i^{[k]}(R) = \left\{\begin{array}{ll} KO_i(R), & k = 0, \\ U_i(R), & k = 1, \\ KSp_i(R), & k = 2, \\ {}_{-1}\!U_i(R), & k = 3. \end{array}\right.
\end{equation}
The notation for the upper index $[k]$ suggests that Grothendieck--Witt groups are $4$-periodic in $k$.
The groups ${}_{\pm 1}\!U_i(R)$ are defined in~\cite{Ka80} and we will not use their definition.

In the sequel we will need the following result.
\begin{theorem}[Bass Fundamental Theorem]\label{bass-ft} Suppose that $2 \in R^*$, then for any $i\geq 1$, $k\in \ZZ/4\ZZ$ there is a split exact sequence of abelian groups
 \[ \xymatrix{ 0 \ar[r] & \GW_i^{[k]}(R) \ar[r] &  \GW_i^{[k]}(R[X, X^{-1}]) \ar[r]  & \GW_{i-1}^{[k-1]}(R) \ar[r] & 0.} \] \end{theorem}
\begin{proof} This is a special case of~\cite[Theorem~9.13]{Sch16} and the remark that follows it. \end{proof}

@@
